### 2024-05-22 thesis/appendix.tex
thesis update
22 May 2024

\subsection{Halving gridsearch}
Gridsearch has been the to go choice for hyperparameters tuning for the past years. However, such method is brute forcing all possible combinations, hence it is highly computationally intensive, especially when it comes to large datasets.
\\
The scikit-learn team addressed this disadvantage of gridsearch by introducing the halving gridsearch method \cite{scikithalvinggridsearch} (2020). Such technique has proved itself to greatly speed up hyperparameter tuning.
The ground concept underlying this method is successive halving. During the first iteration of halving gridsearch, all candidates are trained on a small subset of the training set. Next, we keep only the candidates which performed best and compare  them again on a bigger subset of the training set. As the iterations pass, the surviving candidates will be given more and more training samples. The algorithm stops when we are left with only the best set of hyperparameters.

@@

\subsection{Data normalization}\label{appendix:normalization}
With data normalization we transform the range of the feautures to a standard scale.
Such preprocessing step is essential when employing distance based algorithms like SVM or K-nearest neighbors. The rationale behind it is that by normalizing data we give an uniform weight to each feature in the learning process; in this way we do not favour larger scale features.
Examples of the most popular features scaling are:
\begin{itemize}
    \item Standard scaler= it computes the standard score z of a sample x
    \\
    z=$\frac{x-\mu}{\sigma}$.
    \item MinMax scaler= it maps every data sample to the range 0, 1.
    \\
    $\frac{x-\min(X)}{\max(X)-\min(X)}$
    \item Robust scaler= it scales features using statistics that are robust to outliers.
    Essentially, it subtracts the median and then scales the data according to the interquantile range.
\end{itemize}
Many other data scaling algorithms exists, we refer the reader to a thorough comparison \cite{scikitscalers}.
\\
We conclude this subsection with a custom example that motivates the need of feature scaling \cite{scikitscale_example}. The idea is to compare the results of modelling the data with K-nearest neighbors on the unscaled data against the scaled data. The considered data is the wine recognition dataset \href{https://archive.ics.uci.edu/dataset/109/wine}{https://archive.ics.uci.edu/dataset/109/wine}. The goal for this dataset is recognising from whose cultivator the wine comes based on two features with a completely different scale. The first feature has values in the [0,1000] range while the second feature has values contained in [1,10].
The unscaled and scaled version are compared in figure \ref{fig:feature_scaler_example}.
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/feature_scaler_example.png}
    \caption{importance of feature scaling}
    \label{fig:feature_scaler_example}
\end{figure}
What we can conclude from the image is that, the model trained on scaled data is greatly better than the other.
On the left, we can see that distances between categories are impacted solely by the larger scale feature. Conversely, on the right, we have that the two features contribute equally in determining the neighbors.

@@

\newpage
\section{Source code}\label{src_code}
The whole code for the project is hosted on
\url{https://github.com/luca-pernigo/ThesisKernelMethods}\label{github_repo}.
\\
\begin{itemize}
    \item query/: folder containing Scopus data and scripts to generate bibliometric survey plots in section \ref{literature_review}.
    \item thesis/: folder containing tex files for thesis.
    \item beamer/: folder containing tex files for pptx thesis defence.
    \item experiments: folder containing train, test script for each of the experiments carried out in the thesis alongside with the binary files of the pickled models.
    \item experiments/src/kernel\_quantile\_regression/kqr.py: file implementing our custom kernel quantile regression.
    \item experiments/Data: cleaned data used in our experimental studies.
    \item experiment/plots: folder storing plots resulting from experiments.
    \item experiment/eda/: scripts for exploratory data analysis.
    \item experiments/point/: scripts for experiments point framework.
    \item experiments/train\_test/: scripts for training, testing and generating tables for experiments probabilistic framework.
    \item experiments/utils/: utility functions used in extracting, transforming and cleaning raw data.
    \item extradata/: folder containing data for experiments in section \ref{appendix:quantile_regressor_extensive_comparison}.
    \item requirements.txt: pip freeze of python packages used.
\end{itemize}

>>>>

\subsection{Halving gridsearch}
Gridsearch has been the to go choice for hyperparameters tuning for the past years. However, such method is brute forcing all possible combinations, hence it is highly computationally intensive, especially when it comes to large datasets.
\\
The scikit-learn team addressed this disadvantage of gridsearch by introducing the halving gridsearch method \citeW{scikithalvinggridsearch} (2020). Such technique has proved itself to greatly speed up hyperparameter tuning.
The ground concept underlying this method is successive halving. During the first iteration of halving gridsearch, all candidates are trained on a small subset of the training set. Next, we keep only the candidates which performed best and compare  them again on a bigger subset of the training set. As the iterations pass, the surviving candidates will be given more and more training samples. The algorithm stops when we are left with only the best set of hyperparameters.

@@

\subsection{Data normalization}\label{appendix:normalization}
With data normalization we transform the range of the feautures to a standard scale.
Such preprocessing step is essential when employing distance based algorithms like SVM or K-nearest neighbors. The rationale behind it is that by normalizing data we give an uniform weight to each feature in the learning process; in this way we do not favour larger scale features.
Examples of the most popular features scaling are:
\begin{itemize}
    \item Standard scaler= it computes the standard score z of a sample x
    \\
    z=$\frac{x-\mu}{\sigma}$.
    \item MinMax scaler= it maps every data sample to the range 0, 1.
    \\
    $\frac{x-\min(X)}{\max(X)-\min(X)}$
    \item Robust scaler= it scales features using statistics that are robust to outliers.
    Essentially, it subtracts the median and then scales the data according to the interquantile range.
\end{itemize}
Many other data scaling algorithms exists, we refer the reader to a thorough comparison \citeW{scikitscalers}.
\\
We conclude this subsection with a custom example that motivates the need of feature scaling \citeW{scikitscale_example}. The idea is to compare the results of modelling the data with K-nearest neighbors on the unscaled data against the scaled data. The considered data is the wine recognition dataset \href{https://archive.ics.uci.edu/dataset/109/wine}{https://archive.ics.uci.edu/dataset/109/wine}. The goal for this dataset is recognising from whose cultivator the wine comes based on two features with a completely different scale. The first feature has values in the [0,1000] range while the second feature has values contained in [1,10].
The unscaled and scaled version are compared in figure \ref{fig:feature_scaler_example}.
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/feature_scaler_example.png}
    \caption{importance of feature scaling}
    \label{fig:feature_scaler_example}
\end{figure}
What we can conclude from the image is that, the model trained on scaled data is greatly better than the other.
On the left, we can see that distances between categories are impacted solely by the larger scale feature. Conversely, on the right, we have that the two features contribute equally in determining the neighbors.

@@

\newpage
\section{Source code}\label{src_code}
The whole code for the project is hosted on
\url{https://github.com/luca-pernigo/ThesisKernelMethods}\label{github_repo}.
\\
\begin{itemize}
    \item query/: folder containing Scopus data and scripts to generate bibliometric survey plots in section \ref{literature_review}.
    \item thesis/: folder containing tex files for thesis.
    \item beamer/: folder containing tex files for pptx thesis defence.
    \item experiments: folder containing train, test script for each of the experiments carried out in the thesis alongside with the binary files of the pickled models.
    \item experiments/src/kernel\_quantile\_regression/kqr.py: file implementing our custom kernel quantile regression.
    \item experiments/Data: cleaned data used in our experimental studies.
    \item experiment/plots/: folder storing plots resulting from experiments.
    \item experiment/eda/: scripts for exploratory data analysis.
    \item experiments/point/: scripts for experiments point framework.
    \item experiments/train\_test/: scripts for training, testing and generating tables for experiments probabilistic framework.
    \item experiments/utils/: utility functions used in extracting, transforming and cleaning raw data.
    \item extradata/: folder containing data for experiments in section \ref{appendix:quantile_regressor_extensive_comparison}.
    \item requirements.txt: pip freeze of python packages used.
\end{itemize}

### 2024-05-22 thesis/input_files/energymarket.tex
22 May 2024

\section{Electricity distribution network}
Electricity is generated by power plants which transfer it over the so called transmission level. Then, through the transmission network, this energy is transported at a high voltage over long distances. Finally, the voltage is reduced and the energy is moved into the distribution network. In a nutshell, transmision lines carry electric power from stations to substations while distribution lines carry electricity from substations to load points such as businesses, industries and homes, see figure \ref{electricity_network}.

@@

\subsection{The marketplace}
We can differentiate between two kinds of electricity markets: power pools and power exchanges. In power pools, generators bid the prices at which they are willing to produce at different volumes. Then, the market clearing price (MCP) is determined by intersecting the aggregated supply curve and the estimated demand, left panel figure \ref{fig:pool_vs_echange}. Power pools are created on public initiatives of governments. Conversely, power exchanges are created through private agreements between generators, distributors and traders. This is the model followed by most of european countries.
The MCP in power exchanges is determined by the intersection of the aggregate supply curve and the aggregate demand curve, right panel figure \ref{fig:pool_vs_echange}.
\\
It is worth to mention the two biggest power exchanges: Epex and NordPool. Epex operates throughout continental europe. NordPool covers the nordinc and baltic contries.
\\
Moreover, we can differentiate between two popular types of auctions: uniform-price/marginal and pay as bid/discriminatory. Within the uniform price setting, suppliers offering for less than the clearing price are paid that price. Analogously, consumers bidding more than the clearing price pay that price. On the other hand, in pay as bid auctions, suppliers are paid the exact price they bid for.

@@

% \subsection{Intraday}
The intraday market consists of continous trading 24/7. A trade is executed as soon as a buy and a sell order mathch. Electricity can be traded an hour, half-hour, quarter or also 5 minutes before delivery. This enables market participants to balance their positions in real time, should they need to do so.

\subsection{Price peaks}
Electricity market are characterized by spikes in their spot prices; when this happens, the system price jumps abruptly and then drops back within a very short period.
This spikness follows from the non storability of electricity; that is, electricity has to be consumed as it is produced. Therefore, extreme load fluctuations combined with generation outages or transmission failures can result in price spikes.

@@

\subsection{Nodal and zonal pricing}
Grid configuration and physical limits on electricity lines may lead to congestion. Zonal market clearing pricing (ZMCP) and local marginal price (LMP) are two types of pricing schemes utilized to handle it. The former is adopted in EU contries while the latter is used in the US, figure \ref{fig:eu_zonal} and figure \ref{fig:us_local}.
With zonal market clearing pricing, prices may differ between zones but are the same within the same area.
On the other hand, local marginal price is made up by summing the transmission congestion cost, generation marginal cost and the cost of marginal losses at different buses; buses is where a electicity line or several lines are connected.
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/eu_zonal.png}
    \caption{Zonal pricing EU \cite{eu_zonal}}
    \label{fig:eu_zonal}
\end{figure}

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/us_local.png}
    \caption{Local pricing US \cite{us_local}}
    \label{fig:us_local}
\end{figure}

% pensodino \subsection{Nodal and Zonal Pricing}
\section{Smart grids}
A smart meter is a digital device recording energy consumption in real time and communicating this data to the supplier and the consumer.
Deployment of smart grids and the digitalisation of the electricity network are intended to increase the energy efficiency of our networks.
The reason is that, smart metering and the electricity network digitalisation  will enable system operators to better monitor the network, plan their investments and manage infrastructure. 
All of this motivates for the need of efficient and robust tools to analyze the vast data resulting from the digitalisation of the power sector see \cite{declean,andreadou2016telecommunication}.

>>>>

\section{Electricity distribution network}
Electricity is generated by power plants which transfer it over the so called transmission level. Then, through the transmission network, this energy is transported at a high voltage over long distances. Finally, the voltage is reduced and the energy is moved into the distribution network. In a nutshell, transmission lines carry electric power from stations to substations while distribution lines carry electricity from substations to load points such as businesses, industries and homes, see figure \ref{electricity_network}.

@@

\subsection{The marketplace}
We can differentiate between two kinds of electricity markets: power pools and power exchanges. In power pools, generators bid the prices at which they are willing to produce at different volumes. Then, the market clearing price (MCP) is determined by intersecting the aggregated supply curve and the estimated demand, left panel figure \ref{fig:pool_vs_echange}. Power pools are created on public initiatives of governments. Conversely, power exchanges are created through private agreements between generators, distributors and traders. This is the model followed by most of european countries.
The MCP in power exchanges is determined by the intersection of the aggregate supply curve and the aggregate demand curve, right panel figure \ref{fig:pool_vs_echange}.
\\
It is worth to mention the two biggest power exchanges: Epex and NordPool. Epex operates throughout continental europe. NordPool covers the nordic and baltic countries.
\\
Moreover, we can differentiate between two popular types of auctions: uniform-price/marginal and pay as bid/discriminatory. Within the uniform price setting, suppliers offering for less than the clearing price are paid that price. Analogously, consumers bidding more than the clearing price pay that price. On the other hand, in pay as bid auctions, suppliers are paid the exact price they bid for.

@@

% \subsection{Intraday}
The intraday market consists of continous trading 24/7. A trade is executed as soon as a buy and a sell order match. Electricity can be traded an hour, half-hour, quarter or also 5 minutes before delivery. This enables market participants to balance their positions in real time, should they need to do so.

\subsection{Price peaks}
Electricity markets are characterized by spikes in their spot prices; when this happens, the system price jumps abruptly and then drops back within a very short period.
This spikeness follows from the non storability of electricity; that is, electricity has to be consumed as it is produced. Therefore, extreme load fluctuations combined with generation outages or transmission failures can result in price spikes.

@@

\subsection{Nodal and zonal pricing}
Grid configuration and physical limits on electricity lines may lead to congestion. Zonal market clearing pricing (ZMCP) and local marginal price (LMP) are two types of pricing schemes utilized to handle it. The former is adopted in EU contries while the latter is used in the US, figure \ref{fig:eu_zonal} and figure \ref{fig:us_local}.
With zonal market clearing pricing, prices may differ between zones but are the same within the same area.
On the other hand, local marginal price is made up by summing the transmission congestion cost, generation marginal cost and the cost of marginal losses at different buses; buses is where a electricity line or several lines are connected.
\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/eu_zonal.png}
    \caption[Zonal pricing EU]{Zonal pricing EU \citeW{eu_zonal}}
    \label{fig:eu_zonal}
\end{figure}

\begin{figure}[!h]
    \includegraphics[width=\textwidth]{images/us_local.png}
    \caption[Local pricing US]{Local pricing US \citeW{us_local}}
    \label{fig:us_local}
\end{figure}

% pensodino \subsection{Nodal and Zonal Pricing}
\section{Smart grids}
A smart meter is a digital device recording energy consumption in real time and communicating this data to the supplier and the consumer.
Deployment of smart grids and the digitalisation of the electricity network are intended to increase the energy efficiency of our networks.
The reason is that, smart metering and the electricity network digitalisation  will enable system operators to better monitor the network, plan their investments and manage infrastructure. 
All of this motivates for the need of efficient and robust tools to analyze the vast data resulting from the digitalisation of the power sector, see \cite{declean,andreadou2016telecommunication}.

### 2024-05-22 thesis/input_files/metrics.tex
22 May 2024

Proper evaluation methods guide researchers in choosing the model that best fits their needs; thus, this chapter is dedicated to the most common evaluation metrics adopted by academics in the field of electricity forecasting. Error metrics and measures vary depending on whethter we are concerned with point or probabilistic forecasts. Additionally, note that the latter can take different forms which therefore requires different measures.
\section{Mean absolute error}\label{mae}
Consider the time series with actual values given by $y=(y_{n+1}, y_{n+2},\dots, y_{n+h})$
and its $h$ step ahead point forecast $\hat{y}=(\hat{y}_{n+1}, \hat{y}_{n+2},\dots, \hat{y}_{n+h})$ the mean absolute error (MAE) is defined as
\begin{definition}
    $\mathrm{MAE}(y,\hat{y})=\frac{1}{h}\| y- \hat{y}\|_{1}=\frac{1}{h}\sum\limits_{k=1}^{h}|y_{n+k}-\hat{y}_{n+k}|$
\end{definition}

@@

Mean absolute error and root mean squared error posses the useful property of being expressed in the same units of the data, thus enabling meaningful comparisons.
However, a drawback of such measures is that we cannot use them to compare accuracy between time series which have different magnitudes. For instance, a day ahead error of 1kWh is negligible when considering a daily demand of 100kWn while the same error is considerably big when daily demand is 2kWh. This consideration leads to relative accuracy scores, between those the mean absolute percentage error (MAPE) is by far the most popular.

@@

Where delta is the prediction interval (PI) width, that is $\delta=u_t-l_t$ where $u_t$ is the prediction interval upper threshold and $l_t$ is the prediction interval lower threshold at time step $t$. This score penalizes observations falling outside the prediction interval and rewards narrow prediction intervals.

@@

The continous ranked probability score (CRPS) measures the differece between the estimated cumulative distribution $\hat{F}$ and the empirical cumulative density function (CDF).

>>>>

Proper evaluation methods guide researchers in choosing the model that best fits their needs; thus, this chapter is dedicated to the most common evaluation metrics adopted by academics in the field of electricity forecasting. Error metrics and measures vary depending on whether we are concerned with point or probabilistic forecasts. Additionally, note that the latter can take different forms which therefore requires different measures.
\section{Mean absolute error}\label{mae}
Consider the time series with actual values given by $y=(y_{n+1}, y_{n+2},\dots, y_{n+h})$
and its $h$ step ahead point forecast $\hat{y}=(\hat{y}_{n+1}, \hat{y}_{n+2},\dots, \hat{y}_{n+h})$ the mean absolute error (MAE) is defined as
\begin{definition}
    $\mathrm{MAE}(y,\hat{y})=\frac{1}{h}\| y- \hat{y}\|_{1}=\frac{1}{h}\sum\limits_{k=1}^{h}|y_{n+k}-\hat{y}_{n+k}|$
\end{definition}

@@

Mean absolute error and root mean squared error posses the useful property of being expressed in the same units of the data, thus enabling meaningful comparisons.
However, a drawback of such measures is that we cannot use them to compare accuracy between time series which have different magnitudes. For instance, a day ahead error of 1kWh is negligible when considering a daily demand of 100h while the same error is considerably big when daily demand is 2kWh. This consideration leads to relative accuracy scores, between those the mean absolute percentage error (MAPE) is by far the most popular.

@@

Where $\delta$ is the prediction interval (PI) width, that is $\delta=u_t-l_t$, $u_t$ is the prediction interval upper threshold and $l_t$ is the prediction interval lower threshold at time step $t$. This score penalizes observations falling outside the prediction interval and rewards narrow prediction intervals.

@@

The continous ranked probability score (CRPS) measures the difference between the estimated cumulative distribution $\hat{F}$ and the empirical cumulative density function (CDF).
